Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-05-19 manuscript.tex
Work on time integration section

The principle difficulty of solving \cref{eq:depletion-matrix} is the fact that
the elements of $\mathbf{A}$ are not constant with respect to time. If
$\mathbf{A}$ is constant, \cref{eq:depletion-matrix} admits a closed-form
solution,
\begin{equation}
  \mathbf{n}(t) = \exp \left (\mathbf{A} t \right ) \mathbf{n}_0.
\end{equation}
The exponential of a matrix $\mathbf{X}$ is defined by the power series
expansion
\begin{equation}
    \exp(\mathbf{X}) = \sum\limits_{k=0}^\infty \frac{1}{k!} \mathbf{X}^k
\end{equation}
where $\mathbf{X}^0 = \mathbf{I}$. Most numerical methods for solving
\cref{eq:depletion-matrix} involve two separate components:
\begin{enumerate}
  \item Approximating \cref{eq:depletion-matrix} in a form that involves taking
  one or more matrix exponentials, and
  \item Evaluating the matrix exponential or, equivalently, the action of a
  matrix exponential on a vector.
\end{enumerate}
We will henceforth refer to the first component as \emph{time integration}. Let
us now consider how each of these components are treated in OpenMC.

@@

A variety of numerical methods exist for solving \cref{eq:depletion-matrix}. The
simplest such method, known as the "predictor" method, is to divide the overall
time interval of interest $[0,t]$ into smaller timesteps over which it is
assumed that the burnup matrix is constant. Let $t \in [t_i, t_i + h]$ be one
such timestep. Over the timestep, the solution to \cref{eq:depletion-matrix} can
be written analytically using the matrix exponential
\begin{align}
    \mathbf{A}_i &= \mathbf{A}(\mathbf{n}_i) \\
    \mathbf{n}_{i+1} &= \exp(\mathbf{A}_i h) \mathbf{n}_i
\end{align}
where $\mathbf{n}_i \equiv \mathbf{n}(t_i)$. A series of so-called predictor-corrector
methods that use multiple stages offer improved accuracy over the predictor
method. The simplest of these methods, the CE/CM algorithm, is defined as
\begin{align}
    \mathbf{n}_{i+1/2} &= \exp \left (\frac{h}{2}\mathbf{A}(\mathbf{n}_i) \right) \mathbf{n}_i \\
    \mathbf{n}_{i+1} &= \exp \left(h \mathbf{A}(\mathbf{n}_{i+1/2}) \right) \mathbf{n}_i
\end{align}
Here, the value of $\mathbf{n}$ at the midpoint is estimated using
$\mathbf{A}$ evaluated at the beginning of the timestep. Then,
$\mathbf{A}$ is evaluated using the densities at the midpoint and used to
integrate over the entire timestep.

>>>>

The principle difficulty of solving \cref{eq:depletion-matrix} is the fact that
the elements of $\mathbf{A}$ are not constant with respect to time. If
$\mathbf{A}$ is constant, \cref{eq:depletion-matrix} admits a closed-form
solution,
\begin{equation}
  \label{eq:constant-A}
  \mathbf{n}(t) = \exp \left (\mathbf{A} t \right ) \mathbf{n}_0.
\end{equation}
The exponential of a matrix $\mathbf{X}$ is defined by the power series
expansion
\begin{equation}
    \exp(\mathbf{X}) = \sum\limits_{k=0}^\infty \frac{1}{k!} \mathbf{X}^k
\end{equation}
where $\mathbf{X}^0 = \mathbf{I}$. Most numerical methods for solving
\cref{eq:depletion-matrix} involve two separate components:
\begin{enumerate}
  \item Approximating \cref{eq:depletion-matrix} in a form that involves taking
  one or more matrix exponentials, and
  \item Evaluating the matrix exponential or, equivalently, the action of a
  matrix exponential on a vector.
\end{enumerate}
We will henceforth refer to the first component as \emph{time integration}. Let
us now consider how each of these components are treated in OpenMC.

@@

Because an analytical solution to \cref{eq:depletion-matrix} is not available,
one must resort to using a numerical method that generates approximation
solutions in discrete increments. In general, the problem we need to solve then
is: given $\mathbf{n}_i \equiv \mathbf{n}(t_i)$, what is the value of
$\mathbf{n}_{i+1} \equiv \mathbf{n}(t_{i+1}) = \mathbf{n}(t_i + h)$, with $h$
being the step size? Without loss of generality, we shall choose $t_i=0$ as it
makes the presentation of various methods simpler.

OpenMC doesn't rely on a single time integration method but instead has several
methods available that the user can choose. In the following sections, we
outline the available methods in OpenMC, several of which have not been
described in the literature.

\subsubsection{Predictor Method}

The simplest numerical method for solving \cref{eq:depletion-matrix} over a
timestep is known as the \emph{predictor} method. It proceeds by assuming that
$\mathbf{A}$ is constant over the timestep and assumes the value at the
beginning of the step. The solution then follows directly from
\cref{eq:constant-A}:
\begin{equation}
    \mathbf{n}_{i+1} = \exp\left(h\mathbf{A}(\mathbf{n}_i) \right) \mathbf{n}_i.
\end{equation}
The end-of-step composition is thus ``predicted'' by the reaction rates at the
beginning of the step. This method has the virtue of requiring a single
transport solution and a single matrix exponential but is only first-order
accurate.

\subsubsection{Predictor-Corrector Methods}

A series of \emph{predictor-corrector} methods that use multiple stages offer
improved accuracy over the predictor method. The simplest of these methods is
the CE/CM algorithm, which is the default algorithm used in
MCNP6~\citep{fensin2006tans}. In this method, $\mathbf{A}$ is evaluated at the
beginning of the step and used to deplete the composition over half the step.
Then, $\mathbf{A}$ is reevaluated using the predicted middle-of-step composition
and used to deplete over the full step. The algorithm can be expressed as
\begin{align}
    \mathbf{n}_{i+1/2} &= \exp \left (\frac{h}{2}\mathbf{A}(\mathbf{n}_i) \right) \mathbf{n}_i \\
    \mathbf{n}_{i+1} &= \exp \left(h \mathbf{A}(\mathbf{n}_{i+1/2}) \right) \mathbf{n}_i.
\end{align}
This method requires two transport solutions and two matrix exponential solves
(twice as expensive as the predictor method) and is second-order accurate.

More sophisticated predictor-corrector methods have been introduced by
\citet{isotalo2011ane2,isotalo2011ane3}. In the CE/LI method, the end-of-step
composition is first predicted using $\mathbf{A}(\mathbf{n}_i)$:
\begin{equation}
  \mathbf{n}_{i+1}^p = \exp \left( h\mathbf{A}(\mathbf{n}_i) \right) \mathbf{n}_i
\end{equation}
Then, $\mathbf{n}_{i+1}^p$, is used to reevaluate $\mathbf{A}$. An approximation
of $\mathbf{A}$ that depends on time is formed by linearly interpolating between
$\mathbf{A}(\mathbf{n}_i)$ and $\mathbf{A}\left(\mathbf{n}_{i+1}^p\right)$. This
contitutes a new differential equation for an approximation of $\mathbf{n}$ that
we shall denote $\hat{\mathbf{n}}$,
\begin{equation}
  \frac{d\hat{\mathbf{n}}}{dt} = \left [ \left ( 1 - \frac{t}{h} \right)
  \mathbf{A}(\mathbf{n}_i) + \frac{t}{h} \mathbf{A}\left(\mathbf{n}_{i+1}^p \right) \right ] \hat{\mathbf{n}}
\end{equation}
